---
title: "BA Student Movement Data Cleaning_Submission"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Beach, Hatcher. Coats, Cameron. Sharifi,Mani"
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    code-tools: true 
    code-fold: true
    code-overflow: scroll
    self-contained: true
    #bibliography: winerefs.bib
---
```Goals
Verify the data. If you find it is not correct or consistent, present your approach, and compare/contrast your counts to the existing counts
Operationally define freshman, sophomore, junior, senior for the purposes of this counting exercise
Explain why we had 84 BA majors in Spring 2023 but only 62 graduated in 2024. Is this true? If so, where did the other 22 people go?
Develop a easier and more informative method to present this data
Make it easy to develop and explain to internal and external audience
Make it reproducible so that we can update it each semester
Make it better

Checkpoint 1

This is your first checkpoint for Project 3.  The final project will be delivered to the client on Friday, December 8; however, your team will turn in the final project on Wednesday, November 29.  Presentations to the class will be on November 30, December 5, 7, and 14.  This checkpoint is to ensure you are making adequate progress towards your final deliverable.  This checkpoint should include, at minimum, the following items:

Your problem statement:  A concise paragraph (or bullet items, or a combination) that describes the overall problem you are trying to solve.
This project addresses the inconsistencies in Business Analytics and ISA majors at FSB. In this case, there is a drop from 84 BA majors in Spring of 2023 but only 62 graduated in 2024. The problem is finding where those 22 students went, and how these discrepancies came to life. To solve this problem, we need to define who is a freshman, sophomore, junior, and senior as there is a grey area to how these students are represented. By doing this, we can explain graduation counts more easily and the differences between Spring 2023 and 2024. Our final goal is to deliver insights into student progression, improve data reliability, and enable accessible reporting.

The research questions that outline the specific questions you are going to answer with your analysis
Research questions:
Are current student grade levels accurate? If not, how are they classified and what would be a more accurate way to measure grade level? Credit hours, semesters enrolled, etc.
Why did the number of BA majors decrease from Spring 2023 to 2024? Did these students change majors, graduate early, transfer, etc.?
How can we ensure our analysis is reproducible for upcoming semesters?
What are the most effective ways to present enrollment and graduation data to make it accessible for the client?

A completed Business Value Proposition Canvas
https://www.canva.com/design/DAGVJpL-lKI/Oz2Y_f_C9PH3SL3BPOmbwA/edit 

A list of deliverables that you will provide to the client upon completion of your project
A method of classification(freshman, sophomore, junior, senior) for students 
An overall, cleaned data set based on our method of classification
A complete technical report of what we accomplished
A final presentation slide deck showcasing our major findings for the project


A timeline for completing your project:  This should include the tasks that need to be completed, the person on your team who is responsible for overseeing the completion of the task, the date the task is due to be completed.
Cam will do data cleaning and validation by Monday the 4th
Hatcher will define the problem statement, research questions, BVP, deliverables, and timeline by Monday the 4th
Mani will combine the cleaned data and written content into an RMarkdown for submission for Checkpoint 1 by Monday the 4th
Cam, Hatcher, and Mani will collaborate on a way to classify grade levels by Friday the 9th
All develop and finalize HTML workflow and GitHub for Checkpoint 2 submission on Sunday the 17th
All will collaborate on the technical report and have finished before Friday the 22nd
All will develop the final workflow deliverable after we have finished all the above, so by Friday the 22nd or Monday the 25th
All will collaborate on the final presentation slides and have them complete before Thanksgiving break, Tuesday the 26th
Cam, Hatcher, and Mani will then ensure the above three deliverables are ready to submit on Sunday the 1st and will be ready to present on Tuesday the 3rd



Any other work projects that have been completed.  For example, you may have cleaned and merged your data.  All code and output should be included.  


All of your work should be completed in a single reproducible document such as an RMarkdown or other notebook.  When turning in your document, you must turn in the rendered version in HTML or your team will receive a grade of zero.


Speaker notes
Multiple cohort terms means they might’ve dropped out and came back - take the second one
*SD = major not specified
Count term codes for grade status - issue is students coming in with hours


He will  add ISA classes 




Steps to unmerge and fill excel data 
If you’re working with a large dataset, using VBA (Visual Basic for Applications) can automate this process efficiently. Here's a VBA script to unmerge cells in a selected range and fill each unmerged cell with the original value:
Press Alt + F11 to open the VBA editor in Excel.
Go to Insert > Module to create a new module.
Paste the following code:
vba
Copy code
Sub UnmergeAndFill()
    Dim cell As Range
    Dim mergeArea As Range
    
    For Each cell In Selection
        If cell.MergeCells Then
            Set mergeArea = cell.MergeArea
            cell.MergeCells = False
            mergeArea.Value = cell.Value
        End If
    Next cell
    
    MsgBox "Unmerge and fill complete."
End Sub


Close the VBA editor.
Select the range you want to unmerge and fill, then press Alt + F8 to open the macro window, select UnmergeAndFill, and click Run.
This will unmerge all merged cells in the selected range and fill each cell with the original merged value. It works well for large datasets without needing manual copying and pasting.
```

## Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(DataExplorer, # For explorer and cleaning data by visualizing missing values.
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr, # for nice summary of data
               readxl, #to read Excel files
               networkD3 # to create a Sankey Plot
               ) # For creating dummy variables of categorical variables automatically.)
```

## Load data

### Loading BA Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
BA_Students=read_excel("./BA Major Students.xlsx")
head(BA_Students)
paste("The number of rows in the data is ",nrow(BA_Students))
paste("The number of unique students is ",length(unique(BA_Students$`Student ID`)))
write.csv(BA_Students,"data-clean/BA_Students_cleaned.csv")

```
### Loading IS Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
IS_Students=read_excel("./IS Major Students.xlsx")
head(IS_Students)
paste("The number of rows in the data is ",nrow(IS_Students))
paste("The number of unique students is ",length(unique(IS_Students$`Student ID`)))
write.csv(IS_Students,"data-clean/IS_Students_cleaned.csv")

```

### Loading BA Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
BA_Majors=read_excel("./BA Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
BA_Majors=BA_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`BA Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(BA_Majors)
paste("The number of rows is ",nrow(BA_Majors))
paste("The number of unique students is ",length(unique((BA_Majors$`Student ID`))))

write.csv(BA_Majors,"data-clean/BA_Majors_cleaned.csv")
```
### Loading IS Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
IS_Majors=read_excel("./IS Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
IS_Majors=IS_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(IS_Majors)
paste("The number of rows is ",nrow(IS_Majors))
paste("The number of unique students is ",length(unique((IS_Majors$`Student ID`))))

write.csv(IS_Majors,"data-clean/IS_Majors_cleaned.csv")
```
### Loading BA Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
BA_Grades=read_excel("./BA Major Students - Grades -updated.xlsx")

# Fill in the blank cells created from the format of the data
BA_Grades = BA_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(BA_Grades)
tail(BA_Grades)
paste("The number of unique students is ",length(unique(BA_Grades$`Student ID`)))
num_fresh=BA_Majors %>%
  filter(`Cohort Term`=="202410")%>%
  nrow()
paste("The number of students starting in term 201410 is ", num_fresh)

write.csv(BA_Grades,"data-clean/BAGrades_cleaned.csv")
```
### Loading IS Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
IS_Grades=read_excel("./IS Major Students - Grades -updated.xlsx")

# Fill in the blank cells created from the format of the data
IS_Grades = IS_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(IS_Grades)
tail(IS_Grades)
paste("The number of unique students is ",length(unique(IS_Grades$`Student ID`)))
num_fresh=IS_Majors %>% ##BA
  filter(`Cohort Term`=="202410")%>%
  nrow()
paste("The number of students starting in term 201410 is ", num_fresh)

write.csv(IS_Grades,"data-clean/ISGrades_cleaned.csv")
```
### Loading BA Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
Attributes=read_excel("./BA Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
Attributes = Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(Attributes$`Student ID`)))

head(Attributes)
test=Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(Attributes,"data-clean/BA_Attributes_cleaned.csv")
```

### Loading IS Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
IS_Attributes=read_excel("./IS Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
IS_Attributes = IS_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(IS_Attributes$`Student ID`)))

head(IS_Attributes)
test=IS_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(IS_Attributes,"data-clean/IS_Attributes_cleaned.csv")
```
