---
title: "BA Student Movement Data Cleaning_Submission"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "XX"
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    code-tools: true 
    code-fold: true
    code-overflow: scroll
    self-contained: true
    #bibliography: winerefs.bib
---


## Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(DataExplorer, # For explorer and cleaning data by visualizing missing values.
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr, # for nice summary of data
               readxl, #to read Excel files
               networkD3 # to create a Sankey Plot
               ) # For creating dummy variables of categorical variables automatically.)
```

## Load data

### Loading BA Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
BA_Students=read_excel("./BA Major Students.xlsx")
head(BA_Students)
paste("The number of rows in the data is ",nrow(BA_Students))
paste("The number of unique students is ",length(unique(BA_Students$`Student ID`)))
write.csv(BA_Students,"data-clean/BA_Students_cleaned.csv")

```
### Loading IS Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
IS_Students=read_excel("./IS Major Students.xlsx")
head(IS_Students)
paste("The number of rows in the data is ",nrow(IS_Students))
paste("The number of unique students is ",length(unique(IS_Students$`Student ID`)))
write.csv(IS_Students,"data-clean/IS_Students_cleaned.csv")

```

### Loading BA Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
BA_Majors=read_excel("./BA Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
BA_Majors=BA_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`BA Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(BA_Majors)
paste("The number of rows is ",nrow(BA_Majors))
paste("The number of unique students is ",length(unique((BA_Majors$`Student ID`))))

write.csv(BA_Majors,"data-clean/BA_Majors_cleaned.csv")
```
### Loading IS Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
IS_Majors=read_excel("./IS Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
IS_Majors=IS_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(IS_Majors)
paste("The number of rows is ",nrow(IS_Majors))
paste("The number of unique students is ",length(unique((IS_Majors$`Student ID`))))

write.csv(IS_Majors,"data-clean/IS_Majors_cleaned.csv")
```
### Loading BA Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
BA_Grades=read_excel("./BA Major Students - Grades -updated.xlsx")

# Fill in the blank cells created from the format of the data
BA_Grades = BA_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(BA_Grades)
tail(BA_Grades)
paste("The number of unique students is ",length(unique(BA_Grades$`Student ID`)))
num_fresh=BA_Majors %>%
  filter(`Cohort Term`=="202410")%>%
  nrow()
paste("The number of students starting in term 201410 is ", num_fresh)

write.csv(BA_Grades,"data-clean/BAGrades_cleaned.csv")
```
### Loading IS Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
IS_Grades=read_excel("./IS Major Students - Grades -updated.xlsx")

# Fill in the blank cells created from the format of the data
IS_Grades = IS_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(IS_Grades)
tail(IS_Grades)
paste("The number of unique students is ",length(unique(IS_Grades$`Student ID`)))
num_fresh=IS_Majors %>% ##BA
  filter(`Cohort Term`=="202410")%>%
  nrow()
paste("The number of students starting in term 201410 is ", num_fresh)

write.csv(IS_Grades,"data-clean/ISGrades_cleaned.csv")
```
### Loading BA Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
Attributes=read_excel("./BA Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
Attributes = Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(Attributes$`Student ID`)))

head(Attributes)
test=Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(Attributes,"data-clean/BA_Attributes_cleaned.csv")
```

### Loading IS Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
IS_Attributes=read_excel("./IS Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
IS_Attributes = IS_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(IS_Attributes$`Student ID`)))

head(IS_Attributes)
test=IS_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(IS_Attributes,"data-clean/IS_Attributes_cleaned.csv")
```
